### 2023-12-22 globals/metadata.tex
Initialize mscs
writing the paper
update

\title{Title of the document}
\author{Author of the document}
\date{\today}
\bibliography{globals/papers}

\newcommand{\acknowledge}{
    Acknowledge
}

\newcommand{\makeabstract}{
    \begin{abstract}
        Abstract of the paper.
    \end{abstract}
}

>>>>

\title{A corrected version of Karhumäki’s theorem
       on polynomial $\Nat$-rational series}
\author{
    Lopez Aliaume
}
\date{\today}
\bibliography{globals/papers}

\newcommand{\acknowledge}{
    I would like to thank Gaëtan Douéneau-Tabot
    for mentionning this problem, checking 
    my counter-example, and supporting me in publishing
    this paper.
}

\newcommand{\makeabstract}{
    \begin{abstract}
        We propose a short counter-example to a
        previous result on commutative $\Nat$-rational series
        by Karhumäki, and provide a proof for
        the corrected statement.
        This opens the way to the decidability of
        $\Nat$ polyregular functions 
        among $\Rel$ porylegular functions, which is
        an ongoing research.
    \end{abstract}
}

### 2023-12-22 pnrs.tex
update

@@

There has been recent interest in $\Rel$ and $\Nat$-rational series exhibiting
\emph{polynomial growth} \cite{doueneau2021pebble,bojanczyk2022transducers}.
This computational model, in the counting case, can be traced back to a model
introduced by \textcite{schutzenberger1965finite}. It has been recently proven
that one can decide the growth rate of such functions, and decide whether they
are aperiodic in the case of $\Rel$-rational series \cite{LOPEZ23b}. It was
conjectured in the PhD thesis of one of the authors, that the membership
problem of $\Nat$-rational series among $\Rel$-rational series was decidable
under the assumption of polynomial growth. This conjecture was based on a
result of \textcite{KARH77}, which we will disprove in \cref{sec:c-example}. We
then provide in \cref{sec:proof} a corrected version of the theorem, together
with a valid proof. As a consequence, it does not invalidate the stated
conjecture.

\AP Before stating the result, let us introduce the necessary vocabulary on
rational series, together with concrete examples so that no doubt is allowed as
to their meaning. A $\Rel$-rational series $f \colon \Sigma \to \Rel$ is
\intro{commutative} whenever $f(uv) = f(vu)$ for all $u,v \in \Sigma^*$. A
polynomial $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ is a \intro{$\Rel$-rational
polynomial} if there exists a \reintro{commutative} $\Rel$-rational series $f$
such that $f(w) = P(\seqof{ \card[a]{w} }{a \in \Sigma})$. Similarly, a
\intro{$\Nat$-rational polynomial} if the \reintro{commutative} $\Rel$-rational
series is in fact $\Nat$-rational.

\begin{example}
    With $\Sigma \defined \set{1}$,
    the polynomials $X_1$, and $X_1^2 + 3$ are \kl{$\Nat$-rational polynomial},
    but $- X_1$ is a \kl{$\Rel$-rational polynomial} that is 
    not a \kl{$\Nat$-rational polynomial}.
\end{example}

\begin{example}
    Over a binary alphabet $\Sigma \defined \set{a,b}$,
    the polynomial $P(X_a, X_b) \defined (X_a - X_b)^2$
    is a \kl{$\Rel$-rational polynomial} that is \kl{non-negative},
    but is
    not a \kl{$\Nat$-rational polynomial}.
\end{example}

\AP Let us now give some vocabulary on polynomials with multiple indeterminate
over $\Rel$. 
A \intro{monomial} is a product of variables and integers.
\begin{example}
    $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, 
    but $X + Y$ or $2X^2 + XY$ are not.
\end{example}
A polynomial $P \in \Rel[X_1, \dots, X_k]$ has \intro{non-negative}
if $P(n_1, \dots, n_k) \geq 0$ for all $n_1, \dots, n_k \geq 0$. Beware that we
do not consider negative values, as the numbers $n_i$ will ultimately count the
number of occurrences of a letter in a word.
\begin{example}
    The polynomial $(X - Y)^2$ is \kl{non-negative},
    and so is the polynomial $X^3$, but the polynomial
    $X^2 - 2X$ is not.
\end{example}
In a polynomial $P \in \Rel[X_1, \dots, X_k]$, a monomial 
is a \intro{maximal monomial} if it is a maximal element
for the visibility ordering of polynomials \emph{in $\Rat$}
among monomials with non-zero coefficients in $P$.\footnote{
    One could equivalently define maximal elements by removing the (non-zero)
    multiplicative constant of the monomials, and then
    considering the usual divisibility relation of $\Rel$ polynomials.
}
\begin{example}
    In the polynomial 
    $P(X,Y) \defined X^2 - 2XY + X^2 + X + Y$, the \reintro{maximal monomials}
    are $X^2$, $-2 XY$, and $X^2$.
\end{example}
A \kl{monomial} has \intro{positive coefficient} in a polynomial $P$
if its multiplicative constant is positive.
\begin{example}
    In the polynomial $P(X,Y) \defined (X - Y)^2$, 
    the \kl{positive monomials} are
    $X^2$ and $Y^2$.
\end{example}

We are now ready to restate the theorem
of \citeauthor{KARH77}, in terms of its
class $\CoveredPoly$ of polynomials defined as follows:
\begin{definition}[{\cite[Section 3, page 3]{KARH77}}]
    Let $\vec{X}$ be a finite tuple of indeterminate.
    The class $\CoveredPoly[\vec{X}]$
    is the class of polynomials $P \in \Rel[\vec{X}]$
    that are \kl{non-negative}
    and such that every \kl{maximal monomial} has a \kl{positive coefficient}.
    When the indeterminate are clear from the context, we write
    this class $\CoveredPoly$.
\end{definition}

\begin{theorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
    Let $\Sigma$ be a finite alphabet.
    The set of \kl{$\Nat$-rational polynomials} over $\Sigma$
    coincides with $\CoveredPoly[\seqof{X_a}{a \in \Sigma}]$.
\end{theorem}

@@

    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
    Because $L \cap \set{a,b}^*$ is notoriously not a regular language, we
    conclude that $L$ is not a regular language.
    However, $L = f^{-1}(\set{0})$, which is a regular language
    because $f$ is a $\Nat$-rational series (\textbf{TODO: cite some result?}).
\end{proof}

\begin{corollary}
    The result stated in \cite[Theorem 3.3]{KARH77} is false
    for all polynomials with at least $3$ indeterminate.
\end{corollary}

@@

\begin{lemma}
    The \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
\end{lemma}
\begin{proof}[Proof Idea]
    Let $P$ be a \kl{$\Nat$-rational polynomial}.
    Let $\Sigma$  be the finite alphabet such that
    $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$,
    and $f \colon \Sigma^* \to \Nat$ be the \kl{commutative}
    $\Nat$-rational series computing $P$.
    Because $f$ has polynomial growth, and is a $\Nat$-rational series,
    it is a $\Nat$-polyregular function \textbf{CITE}.
    In particular, 
    there exists a tuple $\vec{x}$ of free variables
    $\MSO$ formulas $\psi_1(\vec{x}), \dots, \psi_n(\vec{x})$,
    and positive coefficients $\alpha_1, \dots, \alpha_n \in \Nat$,
    such that
    \begin{equation*}
        f(w) = \left(\sum_{i = 1}^n \alpha_i \vcount{\psi_i}\right)(w) \quad .
    \end{equation*}

@@

>>>>

\usepackage[left=3cm,right=3cm]{geometry}

@@

Let us first state the result 
of \citeauthor{KARH77}, in terms of its
class $\CoveredPoly$ of polynomials defined as follows:
\begin{definition}[{\cite[Section 3, page 3]{KARH77}}]
    Let $\vec{X}$ be a finite tuple of indeterminate.
    The class $\CoveredPoly[\vec{X}]$
    is the class of polynomials $P \in \Rel[\vec{X}]$
    that are \kl{non-negative}
    and such that every \kl{maximal monomial} has a \kl{positive coefficient}.
    When the indeterminate are clear from the context, we write
    this class $\CoveredPoly$.
\end{definition}

\begin{theorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
    \label{karh:thm}
    Let $\Sigma$ be a finite alphabet.
    The set of \kl{$\Nat$-rational polynomials} over $\Sigma$
    coincides with $\CoveredPoly[\seqof{X_a}{a \in \Sigma}]$.
\end{theorem}

\paragraph*{Context.}
There has been recent interest in $\Rel$ and $\Nat$-rational series exhibiting
\emph{polynomial growth} \cite{doueneau2021pebble,bojanczyk2022transducers}.
This computational model, in the counting case, can be traced back to a model
introduced by \textcite{schutzenberger1965finite}. It has been recently proven
that one can decide the growth rate of such functions, and decide whether they
are aperiodic in the case of $\Rel$-rational series \cite{LOPEZ23b}. It was
conjectured in the PhD thesis of one of the authors, that the membership
problem of $\Nat$-rational series among $\Rel$-rational series was decidable
under the assumption of polynomial growth. This conjecture was based on a
result of \textcite{KARH77}, which we will disprove in \cref{sec:c-example}. We
then provide in \cref{sec:proof} a corrected version of the theorem, together
with a valid proof. As a consequence, it does not invalidate the stated
conjecture.

\AP
\paragraph*{Rational series and regular languages.}
\begin{fact}
    The language $\setof{ w \in \Sigma}{ \card[a]{w} = \card[b]{w}}$
    is not regular.
\end{fact}

\begin{fact}
    regular languages are closed under intersections.
\end{fact}

TODO.

clasical results from \cite{berstel2011noncommutative}.
\begin{fact}
    \label{pre-image-regular:fact}
    The pre-image of a regular language by a $\Nat$-rational series
    is a regular language.
\end{fact}

\begin{fact}
    \label{polynomial-rational-polyreg:fact}
    polynomial + rational = unary polyregular.
\end{fact}

\AP
\paragraph*{Polynomials.}
Before stating the result, let us introduce the necessary vocabulary on
rational series, together with concrete examples so that no doubt is allowed as
to their meaning. A $\Rel$-rational series $f \colon \Sigma \to \Rel$ is
\intro{commutative} whenever $f(uv) = f(vu)$ for all $u,v \in \Sigma^*$. A
polynomial $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ is a \intro{$\Rel$-rational
polynomial} if there exists a \reintro{commutative} $\Rel$-rational series $f$
such that $f(w) = P(\seqof{ \card[a]{w} }{a \in \Sigma})$. Similarly, a
\intro{$\Nat$-rational polynomial} if the \reintro{commutative} $\Rel$-rational
series is in fact $\Nat$-rational.

\begin{example}
    \label{negative-not-nrat:ex}
    With $\Sigma \defined \set{1}$,
    the polynomials $X_1$, and $X_1^2 + 3$ are \kl{$\Nat$-rational polynomial},
    but $- X_1$ is a \kl{$\Rel$-rational polynomial} that is 
    not a \kl{$\Nat$-rational polynomial}.
\end{example}
\begin{proof}
    TODO:
    The polynomial $X_1$ is computed by the rational series,
    the polynomial $X_1^2 + 3$ is computed by,
    and the polynomial
    $- X_1$ cannot be $\Nat$-rational as $\Nat$-rational series
    have non-negative output.
\end{proof}

We saw in \cref{negative-not-nrat:ex}
a simple criterion to check whether a polynomial is $\Nat$-rational,
but being non-negative is not enough, as the following example
illustrates. Note that the proof scheme
of this example will be at the core of \cref{thm:counter-example}.

\begin{example}
    Over a binary alphabet $\Sigma \defined \set{a,b}$,
    the polynomial $P(X_a, X_b) \defined (X_a - X_b)^2$
    is a \kl{$\Rel$-rational polynomial} that is \kl{non-negative},
    but is
    not a \kl{$\Nat$-rational polynomial}.
\end{example}
\begin{proof}
    Assume by contradiction that
    $S$ is a $\Nat$-rational series computing $P$.
    Then, $S^{-1}(\set{0})$ should be a regular language
    (\cref{pre-image-regular:fact}),
    but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
    is not a regular languge.
\end{proof}

\AP Let us now give some vocabulary on polynomials with multiple indeterminate
over $\Rel$. 
A \intro{monomial} is a product of variables and integers.
\begin{example}
    $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, 
    but $X + Y$ or $2X^2 + XY$ are not.
\end{example}
A polynomial $P \in \Rel[X_1, \dots, X_k]$ is \intro{non-negative}
if $P(n_1, \dots, n_k) \geq 0$ for all $n_1, \dots, n_k \geq 0$. Beware that we
do not consider negative values, as the numbers $n_i$ will ultimately count the
number of occurrences of a letter in a word.
\begin{example}
    The polynomial $(X - Y)^2$ is \kl{non-negative},
    and so is the polynomial $X^3$, but the polynomial
    $X^2 - 2X$ is not.
\end{example}
In a polynomial $P \in \Rel[X_1, \dots, X_k]$, a monomial 
is a \intro{maximal monomial} if it is a maximal element
for the visibility ordering of polynomials \emph{in $\Rat$}
among monomials with non-zero coefficients in $P$.\footnote{
    One could equivalently define maximal elements by removing the (non-zero)
    multiplicative constant of the monomials, and then
    considering the usual divisibility relation of $\Rel$ polynomials.
}
\begin{example}
    In the polynomial 
    $P(X,Y) \defined X^2 - 2XY + X^2 + X + Y$, the \reintro{maximal monomials}
    are $X^2$, $-2 XY$, and $X^2$.
\end{example}
A \kl{monomial} has \intro{positive coefficient} in a polynomial $P$
if its multiplicative constant is positive.
\begin{example}
    In the polynomial $P(X,Y) \defined (X - Y)^2$, 
    the \kl{positive monomials} are
    $X^2$ and $Y^2$.
\end{example}

@@

    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
    Because $L \cap \set{a,b}^*$ is notoriously not a regular language, we
    conclude that $L$ is not a regular language.
    However, $L = f^{-1}(\set{0})$, which is a regular language
    because $f$ is a $\Nat$-rational series
    (\cref{pre-image-regular:fact}).
\end{proof}

\begin{corollary}
    The result stated in \cite[Theorem 3.3]{KARH77}, restated
    in \cref{karh:thm}, is false
    for all polynomials with at least $3$ indeterminate.
\end{corollary}

@@

\begin{lemma}
    The \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
\end{lemma}
\begin{proof}[Proof Idea]
    Let $P$ be a \kl{$\Nat$-rational polynomial}.
    Let $\Sigma$  be the finite alphabet such that
    $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$,
    and $f \colon \Sigma^* \to \Nat$ be the \kl{commutative}
    $\Nat$-rational series computing $P$.
    Because $f$ has polynomial growth, and is a $\Nat$-rational series,
    it is a $\Nat$-polyregular function 
    (\cref{polynomial-rational-polyreg:fact}).
    In particular, 
    there exists a tuple $\vec{x}$ of free variables
    $\MSO$ formulas $\psi_1(\vec{x}), \dots, \psi_n(\vec{x})$,
    and positive coefficients $\alpha_1, \dots, \alpha_n \in \Nat$,
    such that
    \begin{equation*}
        f(w) = \left(\sum_{i = 1}^n \alpha_i \vcount{\psi_i}\right)(w) \quad .
    \end{equation*}

@@

\begin{itemize}
    \item Actually, we have an algorithm that allows
        us to \emph{decide} if a polynomial is $\Nat$-rational.
    \item Actually, we obtain a characterisation of
        commutative 
        $\Nat$-rational series of polynomial growth
        among $\Rel$-rational series.
    \item The question of $\Nat$-polyregular functions
        among $\Rel$-polyregular functions remains open
        because of the non-commutativity of the input.
    \item The question of $\Nat$-rational series (commutative)
        among $\Rel$-rational series remains open.
\end{itemize}
